---
title: "Spatial Analysis"
subtitle: "FSH 507 Fall 2019"
author: "Elizabeth Ng"
date: " `r Sys.Date()`"
output:
  ioslides_presentation: default
  powerpoint_presentation: default
  html_document: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(gapminder)
```

# Motivation

- An example going back and forth between ArcGIS and R
- Reproducible workflows!


# Overview & Objectives

- Introduction to the sf package
- Reading and creating shapefiles
- Applying transformations and spatial operations
- Plotting using sf and ggplot2


# Brief background on spatial analysis in R


## R packages for spatial analysis

- Some tools early on
- `rgdal` released in 2003--import from more geographic data formats
- `sp` released in 2005--creates spatial objects with classes and generic methods for points, lines, polygons, grids, and attribute data
-  But still lacked ability to do geometric operations
- `raster` released in 2010--support for raster operations and more


## R packages linking to GIS software

- `GRASS`, `spgrass6`, and `rgrass7`
- `RSAGA`, `RPyGeo`, and `RQGIS`





# Useful links
https://georepository.com/crs_32619/WGS-84-UTM-zone-19N.html
https://spatialreference.org/ref/epsg/2033/

# R Markdown

## Getting Started

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown for presentations see <https://bookdown.org/yihui/rmarkdown/presentations.html>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
pressure %>% 
  ggplot(aes(temperature, pressure)) + 
  geom_point()

```


# Materials



