---
title: "Spatial Analysis"
subtitle: "FSH 507 Fall 2019"
author: "Elizabeth Ng"
date: " `r Sys.Date()`"
output:
  ioslides_presentation: default
  powerpoint_presentation: default
  html_document: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library("tidyverse")
library("sf")
```

# Motivation

- An example going back and forth between ArcGIS and R
- Reproducible workflows!


# Overview & Objectives

- Introduction to the sf package
- Reading and creating shapefiles
- Applying transformations and spatial operations
- Plotting using sf and ggplot2


# Brief background on spatial analysis in R


## R packages for spatial analysis

- Some tools early on
- `rgdal` released in 2003--import from more geographic data formats
- `sp` released in 2005--creates spatial objects with classes and generic methods for points, lines, polygons, grids, and attribute data
-  But still lacked ability to do geometric operations
- `raster` released in 2010--support for raster operations and more


## R packages linking to GIS software

- `GRASS`, `spgrass6`, and `rgrass7`
- `RSAGA`, `RPyGeo`, and `RQGIS`

## Visualization

- `ggplot2`
- `ggmap`
- `rasterVis`
- `tmap`
- `leaflet`
- `mapview`

## Data models

- *vector data model* uses points, lines, and polygons
- *raster data model* divides surface into cells of constant size
- today we'll be dealing with vector data using the `sf` package

## Vector data model

- based on points in a coordinate reference system (CRS)
- stand alone points
- linked together into lines and polygons
- geographic CRS (lon/lat) gives distance from the origin in degrees
- projected CRS (easting/northing) gives distance from the origin in km



# `sf` Package

## Simple Features for R

- [Simple Features package](https://r-spatial.github.io/sf/index.html)
- Based on [simple features ISO standards](https://en.wikipedia.org/wiki/Simple_Features) that are not R specific
- can replace `sp`, `rdgal` for read/write, and `rgeos` for geometric operations

## Basic structure

- tibble for attribute data
- geometry is a list column

```{r}

```





# Resources

## Useful links
https://georepository.com/crs_32619/WGS-84-UTM-zone-19N.html
https://spatialreference.org/ref/epsg/2033/

## Useful links
[Spatial Data Science](https://keen-swartz-3146c4.netlify.com/)



